---
title: "On the inference of positive and negative interactions and their relation to abundance"
author: "Andrew J. Rominger"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    number_sections: no
csl: nature.csl
bibliography: RarePlusComMinus.bib
---

A recent paper infered species-species interactions from spatially replicated abundance data and further tied those infered interactions to abundances [@calatayud2019]. Specifically, the authors found that infered positive interactions were disproportionately associated with rare species, a result they interpreted as rare species being stabilized by facilitative interactions.


```{r setup}
library(knitr)
opts_chunk$set(echo = TRUE, cache = TRUE, 
               fig.width = 4, fig.height = 4, fig.align = 'center')

library(pika)
library(RarePlusComMinus)
library(parallel)
library(MASS)
library(socorro)

# threading defaults
nthrd <- detectCores()
nthrd <- ifelse(round(nthrd * 0.8) >= nthrd - 1, nthrd - 1, round(nthrd * 0.8))
if(nthrd < 1) nthrd <- 1

# plotting defaults
parArgs <- list(mar = c(3, 3, 0, 0) + 0.5, mgp = c(1.75, 0.25, 0), tcl = -0.2)
cexDefault <- 1.4
lwdDefault <- 2
```


First we explore the shape of the spatial abundance patters in the real data provided by ref. @calatayud2019.

```{r obs_data}
# load data from paper
data('abundace_matrices')

# clean it
obsdat <- lapply(abun.mat, function(x) {
    x <- ceiling(x)
    x <- x[rowSums(x) > 0, colSums(x) > 0]
    
    return(x)
})
```

```{r ssad_comp}
summStats <- mclapply(1:length(obsdat), mc.cores = nthrd, FUN = function(i) {
    x <- obsdat[[i]]
    o <- sapply(1:ncol(x), function(j) {
        # sometimes optimization fails, so catch that and return NA
        params <- try(fitdistr(x[, j], 'negative binomial'))
        if(class(params) == 'try-error') {
            params <- c(size = NA, mu = NA, loglik = NA)
        } else {
            params <- c(params$estimate, loglik = params$loglik)
        }
        
        nocc <- sum(x[, j] > 0)
        abund <- sum(x[, j])
        
        return(c(nsite = nrow(x), nspp = ncol(x), nocc = nocc, abund = abund, J = sum(x),
                 params))
    })
})

summStats <- data.frame(study = rep(names(obsdat), sapply(summStats, ncol)),
                        t(do.call(cbind, summStats)))

# remove failed optims
summStats <- summStats[!is.na(summStats$size), ]

# looks like an optim bug around size = 100, remove it
summStats[round(summStats$size) == 100, c('size', 'mu', 'loglik')] <- NA
```

```{r ssad_plotting, eval = FALSE}
par(parArgs)
plot(summStats$nsite, summStats$J, log = 'xy')
plot(summStats$J, summStats$nspp, log = 'xy')

# plot(summStats$nocc / summStats$nsite, summStats$mu, log = 'xy')
# plot(summStats$nocc / summStats$nsite, summStats$size, log = 'xy')
# plot(summStats$nocc, summStats$mu, log = 'xy')
# plot(summStats$nocc, summStats$size, log = 'xy')
# plot(summStats$mu, summStats$size, log = 'xy')
# 
# plot(summStats$abund / summStats$J, summStats$mu, log = 'xy')
# plot(summStats$abund / summStats$J, summStats$size, log = 'xy')
# plot(summStats$abund / summStats$nsite, summStats$mu, log = 'xy')
# plot(summStats$abund / summStats$nsite, summStats$size, log = 'xy')
# plot(summStats$abund, summStats$mu, log = 'xy')
# plot(summStats$abund, summStats$size, log = 'xy')
# 
# plot(summStats$abund / summStats$J, summStats$nocc / summStats$nsite, log = 'xy')
# plot(summStats$abund, summStats$nocc, log = 'xy')
# 
# 
# 
# plot(summStats$nspp, summStats$mu, log = 'xy')
# plot(summStats$nspp, summStats$size, log = 'xy')
# plot(summStats$nspp, summStats$nocc / summStats$nsite, log = 'xy')
```


```{r obs_plusMinus_comp}
commStats <- mclapply(obsdat, mc.cores = nthrd,
                      FUN = function(x) unlist(plusMinus(x)))

commStats <- as.data.frame(do.call(rbind, commStats))

commStats[is.na(commStats$pos.rho.rho) | is.na(commStats$neg.rho.rho), ] <- NA
sum(!is.na(commStats$pos.wm))
```

```{r obs_plusMinus_plot}
# ----
# helper function for making fancy histograms
specialHist <- function(x, breaks, col, add = FALSE, ...) {
    h <- hist(x, breaks, plot = FALSE)

    if(!add) plot(range(h$breaks), range(h$counts), type = 'n', ...)

    rect(xleft = h$breaks[-length(h$breaks)], xright = h$breaks[-1], ybottom = 0, ytop = h$counts,
         col = col)
}


# ----
# plotting

split.screen(c(1, 2))

# correlation differences
screen(1)

par(parArgs)
par(mar = parArgs$mar + c(0, 0, 0, -0.5))
specialHist(commStats$pos.rho.rho - commStats$neg.rho.rho, xlim = c(-2, 2),
            breaks = seq(-2, 2, by = 1/5), col = 'gray90',
            xlab = '', ylab = 'Number of assemblages')
mtext('Abundance-species degree differences\n(positive - negative)',
      side = 1, line = 2.5)
abline(v = 0, col = 'red', lty = 2, lwd = 2)


# raw correlations
split.screen(matrix(c(0.475 + 0.02, 0.975 + 0.02, 0.475, 0.975), nrow = 1), erase = FALSE)
screen(3, new = FALSE)

par(parArgs)
par(cex = 0.75)
par(mgp = par('mgp') * par('cex'))

specialHist(commStats$pos.rho.rho, xlim = c(-1, 1), ylim = c(0, 80),
            breaks = seq(-1, 1, by = 1/10), col = hsv(0.65, alpha = 0.5),
            xlab = expression("Spearman's"~rho),
            ylab = '')
specialHist(commStats$neg.rho.rho,
            breaks = seq(-1, 1, by = 1/10), col = hsv(0, alpha = 0.5),
            add = TRUE)


# mean differences
screen(2)
par(parArgs)
par(mar = parArgs$mar + c(0, -1, 0, +0.5))

specialHist(commStats$pos.wm - commStats$neg.wm, xlim = c(-0.3, 0.3),
            breaks = seq(-0.3, 0.3, by = 0.1/3), col = 'gray90',
            xlab = '', ylab = '')
mtext('Abundance differences\n(positive - negative)',
      side = 1, line = 2.5)
abline(v = 0, col = 'red', lty = 2, lwd = 2)


# raw means
split.screen(matrix(c(0.475 - 0.04, 0.975 - 0.04, 0.475, 0.975), nrow = 1), erase = FALSE)
screen(4, new = FALSE)

par(parArgs)
par(cex = 0.75)
par(mgp = par('mgp') * par('cex'))

specialHist(commStats$pos.wm, xlim = c(0, 0.3),
            breaks = seq(0, 0.3, by = 0.1/6), col = hsv(0.65, alpha = 0.5),
            xlab = 'Weighted mean abund.',
            ylab = '')
specialHist(commStats$neg.wm,
            breaks = seq(0, 0.3, by = 0.1/6), col = hsv(0, alpha = 0.5),
            add = TRUE)
```


```{r test_plot}
plot(1)
plot(1:10)
```

```{r sim_plusMinus_comp, eval = FALSE}
S <- 150
b <- 0.001
nsite <- 25

x <- rfish(S, b)
y <- rnbinom(nsite * length(x), size = 0.1, mu = rep(x, nsite) / nsite)
y <- t(matrix(y, ncol = nsite))
y <- y[rowSums(y) > 0, ]
```

# References

