---
title: "On the inference of positive and negative species interactions and their relation to abundance"
author: "Andrew J. Rominger"
header-includes:
   - \usepackage{xr}
   - \externaldocument{RarePlusComMinus_supp}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: no
csl: nature.csl
bibliography: RarePlusComMinus.bib
---



```{r setup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE, message = FALSE, warning = FALSE)

# function for cross-cache lazy loading
findChunk <- function(key, otherDoc) {
    f <- file.path(paste(otherDoc, 'cache', sep = '_'), 'latex')
    
    allChunk <- list.files(f, pattern = '.RData')
    
    thisChunk <- gsub('\\.RData', '', grep(paste0(key, '_'), allChunk, value = TRUE))
    thisChunk <- file.path(f, thisChunk)
    
    return(thisChunk)
}

# function for cross-doc figure referencing
refOtherFig <- function(x, otherDoc, prefix = '') {
    
}

# load packages
pkgs <- readLines('RarePlusComMinus_supp_cache/latex/__packages')
for(p in pkgs) library(p, character.only = TRUE)

# load plotting defaults
lazyLoad(findChunk('plot_defaults', 'RarePlusComMinus_supp'))
figW <- figH <- 3.75
knitr::opts_chunk$set(fig.width = figW, fig.height = figH, fig.align = 'center')

# load threading defaults
lazyLoad(findChunk('threading', 'RarePlusComMinus_supp'))
```



Why do rare species persist in ecosystems? Rare species seem to be at a disadvantage by pure probabilistic odds [@mcgill2005] and perhaps also from poorly adapted species-environment and species-species interactions [@hutchinson1961], though negative density-dependence may help rare species persist [@leigh2004; @yenni2012]. The question of rarity and persistence thus remains unresolved.  In a recent paper, Calatayuda et al. [@calatayud2019] (CEA) inferred species-species interaction networks from spatially replicated abundance data across many taxa and environments.  CEA found that rare species were more associated with positive interactions compared to common species being more associated with negative interactions, indicating that positive interactions, such as facilitation, may help rare species persist [@calatayud2019]. However, the use of abundance and co-occurrence data to infer species interactions is difficult and often inaccurate [@freilich2018; @carr2019; @rajala2019]. This issue arises in no small part because the underlying null models used to infer interactions themselves are known to have type I and II error problems in real world applications [@ulrich2010; @ladau2008]. Here, I show that the finding of an association between abundance and interaction type as reported by CEA can be explained by statistical artifacts in the inference of species interactions from abundance data. These artifacts arise because of spatial clustering in species abundances. It would therefore not be supported to assign biological interpretations to associations between interaction types and abundances until more data can be brought to bear on the subject of interaction types and the persistence of rare species.

When interaction networks are inferred from spatially replicated abundance data, species-species co-occurrences are quantified by a metric (e.g., CEA use Schoener similarity [@schoener1968]) and then a null model is used to assess whether these co-occurrence metrics deviate substantially enough from null expectations to suggest a non-random interaction, either in the positive direction (suggesting positive interactions) or the negative direction (suggesting negative interactions). However, if abundances are driven by processes, both probabilistic or deterministic, other than species-species interactions, these null models may not reveal true interactions, but rather artifacts of other processes.  Species abundances are not evenly distributed within species nor across space (often referred to as spatial clustering) [@mcgill2003; @engen2008; @zillio2010; @harte2011; @connolly2017]. This ubiquitous pattern can be accounted for by purely probabilistic processes from neutral birth-death-immigration [@kendall1949, @hubbell2001] to mechanistically agnostic statistical-mechanical properties of large assemblages [@harte2011]. Thus, the simple observation of uneven abundances does by itself indicate the influence of deterministic species interactions. The data compiled by CEA [@calatayud2019] indeed confirm the ubiquity of uneven species abundances (Supplementary Figs. \ref{fig:ssad} and \ref{fig:sadShape}). 

In Figure \ref{fig:plusMinus} I first reproduce key results from CEA's Figure 2(B-C). Then to evaluate whether these results can be produced simply from spatial clustering alone I simulate purely random data that match the unevenness of abundance found in the observed data but contain absolutely no species interactions. These random data are simulated as follows:

1) The number of species $S$, number of sites $M$, and shape of the best fitting species abundance distribution (SAD) are sampled (with replacement) from the observed data
2) $S$ species abundances $x_i \ldots x_S$ are sampled from the SAD
3) For each $x_i$, within-species counts are distributed across the $M$ sites according to a spatial species abundance distribution (SSAD) that is either negative binomial (in the case of spatial clustering) or Poisson (in the case of spatial evenness)
4) The resulting simulated site by species matrix is fed through the same analytically pipeline (described in CEA) as the observed data to infer positive and negative interactions.

All analyses are carried out in R [@rcore] and can be fully reproduced by installing the R package accompanying this paper, as detailed in the supplement.

In the case of a Poisson SSAD the one parameter (the mean) is fully specified by the average site-level abundance of a given species. In the case of a negative binomial SSAD, the mean parameter is again specified by the site-level average, but the size or clustering parameter $k$ is not fully specified.  To capture the rough features of the data, I sample $k$ from a linear relationship (with noise) between the maximum likelihood estimates of $k$ and the relative abundance of each species (Supplementary Fig. \ref{fig:ssad}).

```{r fig_plusMinus_prep, include = FALSE}
# load plotting funs
lazyLoad(findChunk('plusMinus_plotFuns', 'RarePlusComMinus_supp'))

# load all data to plot
lazyLoad(findChunk('obs_plusMinus_comp', 'RarePlusComMinus_supp'))
lazyLoad(findChunk('sim_plusMinus_comp', 'RarePlusComMinus_supp'))
lazyLoad(findChunk('sim_pois_plusMinus_comp', 'RarePlusComMinus_supp'))
# lazyLoad(findChunk('sim_poisUnif_plusMinus_comp', 'RarePlusComMinus_supp'))
```

```{r fig_plusMinus, fig.width = figW * 2, fig.height = figH * 2.4, fig.cap = 'Distributions of correlations between network centrality (i.e. species degree) and abundance (left panels) and distributions of weighted mean abundances (right panels). The main figures show the differences between positive and negative interaction networks, while the inset figures show the sepparate distributions for each network. The results of CEA Figure 2(B-C) are reproduced here in panels A-B; panels C-D show data simulated with a negative binomial SSAD and no species interactions; panels E-F show data simulated with a Poisson SSAD and no species interactions. \\label{fig:plusMinus}'}
wmmax <- ceiling(max(simPMData$pos.wm, simPMData$neg.wm, na.rm = TRUE) * 9) / 3
xless <- 0.48
yextra <- 0.55

foo <- split.screen(c(3, 1))

screen(1)
fig2bc(commStats, breaksRho = seq(-2, 2, by = 1/5), 
       breaksWM = seq(-wmmax, wmmax, by = 0.1/3), addxlab = FALSE, 
       figLabs = c('A', 'B'), insetxprop = xless, insetyprop = yextra)
mtext('Observed', side = 4, outer = TRUE)

screen(2)
fig2bc(simPMData, breaksRho = seq(-2, 2, by = 1/5),
       breaksWM = seq(-wmmax, wmmax, by = 0.1/3), addxlab = FALSE,
       figLabs = c('C', 'D'), insetxprop = xless, insetyprop = yextra)

screen(3)
fig2bc(simPMDataPois, breaksRho = seq(-2, 2, by = 1/5),
       breaksWM = seq(-wmmax, wmmax, by = 0.1/3), addxlab = TRUE,
       figLabs = c('E', 'F'), insetxprop = xless, insetyprop = yextra)

foo <- close.screen(all.screens = TRUE)
```

Figure \ref{fig:plusMinus} A--D shows that with a negative binomial SSAD, simulated data closely match observed findings: the correlation between abundance and species' network degree skews more negative in positive interaction networks (i.e. more rare species are more highly connected in positive interaction networks), and positive interaction networks tend to contain more rare species than negative networks. This correspondence between real and simulated patterns largely disappears when we instead use a Poisson SSAD, highlighting the importance of spatial aggregation in driving the spurious results. 


My findings do not depend on simulating SAD and SSAD shapes from the data: in Supplementary Figure \ref{fig:simpleSim} I show that the spurious relationship between abundance and interaction type occurs even when simulating data from just one arbitrary SAD function with the one arbitrary spatially clustered SSAD for all species.  In this simulation, again, replacing the spatially clustered SSAD with a Poisson SSAD breaks the spurious association as in Figure \ref{fig:plusMinus} (E-F).

Why do negative binomial SSADs reproduce the results while Poisson SSADs fail to? The null model algorithm used here and in CEA fixes row and column marginals, but within any given species, the way its total abundance is allocated across sites by the null model has a potentially large combinatorial space to explore. I compare known SSADs to their permuted counterpart in Figure \ref{fig:ssadPerm} and find that the null model transforms negative binomial SSADs to a more Poisson shape, while leaving Poisson SSADs probabilistically unchanged. Specifically, when starting with a negative binomial SSAD, the null model inflates the number of sites individuals are allocated to (more similarly to a Poisson SSAD) and increases the inferred $k$ parameter, indicating less spatial clustering in the permuted matrices compared to their non-permuted, negative binomial starting points.

```{r ssadPerm_prep, include = FALSE}
lazyLoad(findChunk('ssad_perm_comp', 'RarePlusComMinus_supp'))
```

```{r ssadPerm_plot, fig.width = figW * 1.75, fig.cap = "Comparison of SSAD statistics for true and permuted site by species matrices. Colors correspond to the true, un-permuted SSAD. Panel (A) shows how permutation affects number of occupied sites. Panel (B) shows how permutation affects maximum likelihood estimates of the clustering parameter $k$ (B). Points are semi-transparent to help display density. Lines are 1:1 lines. \\label{fig:ssadPerm}"}
nbCol <- hsv(0.53, 1, 0.8)
poCol <- hsv(0.1, 1, 0.95)

layout(matrix(c(1, 1, 2, 3), nrow = 2, byrow = TRUE), heights = c(1, 5))

lmar <- parArgs$mar
lmar[c(1, 3)] <- 0
par(mar = lmar)

plot(1, type = 'n', axes = FALSE, xlab = '', ylab = '')
legend(1, 1, legend = c('Negative binomial', 'Poisson'), pch = 21, pt.cex = 2,
       col = c(nbCol, poCol), pt.bg = colAlpha(c(nbCol, poCol), 0.4),
       bty = 'n', horiz = TRUE, xjust = 0.5)

par(parArgs)
plot(simPoPerm$nocc, simPoPerm$null_nocc, pch = 21,
     col = poCol, bg = colAlpha(poCol, 0.1),
     xlim = c(1, nsite), ylim = c(1, nsite),
     xlab = 'True num. occupied sites',
     ylab = 'Permuted num. occupied sites')

points(simNBPerm$nocc, simNBPerm$null_nocc,
       pch = 21, col = nbCol, bg = colAlpha(nbCol, 0.1))

abline(0, 1, col = 'black', lwd = 2)


foo <- rbind(simNBPerm, simPoPerm)
foo <- foo[is.finite(foo$size) & is.finite(foo$null_size), ]

par(parArgs)
plot(simPoPerm$size, simPoPerm$null_size, log = 'xy', xaxt = 'n', yaxt = 'n',
     pch = 21, col = poCol, bg = colAlpha(poCol, 0.1),
     xlim = range(foo[, c('size', 'null_size')]),
     ylim = range(foo[, c('size', 'null_size')]),
     xlab = expression('True'~k),
     ylab = expression('Permuted'~k))
logAxis(1:2, expLab = TRUE)

points(simNBPerm$size, simNBPerm$null_size,
       pch = 21, col = nbCol, bg = colAlpha(nbCol, 0.1))

abline(0, 1, lwd = 1.5)
```

```{r, mathEG_prep, include = FALSE}
lazyLoad(findChunk('mathmat_fun', 'RarePlusComMinus_supp'))
lazyLoad(findChunk('mathEG_setup', 'RarePlusComMinus_supp'))
lazyLoad(findChunk('mathEG_rare', 'RarePlusComMinus_supp'))
lazyLoad(findChunk('mathEG_comm', 'RarePlusComMinus_supp'))
lazyLoad(findChunk('mathEG_commMax', 'RarePlusComMinus_supp'))

formatNice <- function(x) {
    o <- formatC(x, format = 'e', digits = 2)
    o <- gsub('e', ' \\\\times 10^{', o)
    o <- gsub('10\\^\\{-0', '10^\\{-', o)
    o <- gsub('10\\^\\{0', '10^\\{', o)
    
    return(paste0(o, '}'))
}
```

At a mathematical level, clustered SSADs, compared to spatially even SSADs, actually increase the probability that rare species will appear aggregated with each other and common species will appear repelled.  Consider for example two rare species, one with a single individual and the other with abundance `r Nrare`, distributed across `r nsite` sites. Their Schoener similarity is maximized when all individuals occur at the same site, such as this site by species matrix
$$
X_{rare} = `r mathmat(rarePair)`
$$
If we define $Q(x_i; \mu = `r Nrare / nsite`)$ as the probability of observing $x_i$ individuals in site $i$ given an SSAD with mean parameter $\mu$, then the probability of the above configuration is $P(X_{rare}) = Q(`r Nrare`; \mu = `r Nrare / nsite`) \left(Q(0; \mu = `r Nrare / nsite`)^{`r nsite - 1`}\right)$. Under a negative binomial SSAD with $k = `r k`$, $P(X_{rare}) = `r formatNice(nbRareP)`$ whereas under a Poisson SSAD $P(X_{rare}) = `r formatNice(poRareP)`$. 

Conversely, for two common species, say each with abundance `r Ncomm`, an example configuration that *minimizes* their Schoener similarity would be
$$
Y_{min} = `r mathmat(commPair)`
$$
We calculate the probability of any such scenario where no abundances overlap as $P(Y_{min}) = `r nsite - 1` \left(\left(Q(`r Ncomm`; \mu = `r Ncomm / nsite`) Q(0; \mu = `r Ncomm / nsite`)^{`r nsite - 1`}\right)^2\right)$. With a negative binomial SSAD with $k = `r k`$, $P(Y_{min}) = `r formatNice(nbCommP)`$ whereas with a Poisson SSAD $P(Y_{min}) = `r formatNice(poCommP)`$.

We contrast this with a configuration that would maximize the Schoener similarity between these two common species:
$$
Y_{max} = `r mathmat(commPairMax)`
$$
The probability of this configuration is $P(Y_{max}) = Q(`r commPairMax[1, 1]`; \mu = `r Ncomm / nsite`)^{`r 2 * nsite`}$. For the same negative binomial $P(Y_{max}) = `r formatNice(nbCommPMax)`$, and for the Poisson $P(Y_{max}) = `r formatNice(poCommPMax)`$.

Thus a spatially clustered SSAD, compared to a spatially even SSAD, gives more probability to configurations where rare species appear aggregated and common species appear over-dispersed. Because the null model algorithm permutes site by species matrices to resemble more spatially even SSADs this probabilistic difference between spatially clustered versus even SSADs accounts for the prevalence of rare species in positive interaction networks and common species in negative interaction networks. 

```{r}
# function to calculate percent of species in networks
percNet <- function(x, type) {
    round(mean(100 * x[, paste(type, 'v', sep = '.')] / x$all.v, na.rm = TRUE))
}
```

Great caution should be used when inferring species interactions from abundance data.  More fundamentally than the spurious association of abundance with interaction type, my analysis shows that statistically significant species interactions are inferred from data simulated without any interactions.  In data simulated with a negative binomial SSAD, on average `r percNet(simPMData, 'pos')`% of species were placed in positive interaction networks and `r percNet(simPMData, 'neg')`% in negative interaction networks with a significance cutoff of $\alpha = 0.05$.  With the Poisson SSAD these simulated numbers were `r percNet(simPMDataPois, 'pos')`% for positive networks and `r percNet(simPMDataPois, 'neg')`% for negative networks.  For the observed data, on average `r percNet(commStats, 'pos')`% of species were placed in positive interaction networks and `r percNet(commStats, 'neg')`% in negative interaction networks. Given that these inference methods detect a plethora of interactions where there are none, we cannot be confident in their interpretation.


\clearpage

# References

